Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2015-04-02 paint-contour-tree.tex
Minor edits

\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
}
\date{}

@@

Let us interpret the running time. 
For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
A better upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
This is clearly an improvement when the tree is short and fat, but it can be even better. 
A calculation yields that for any decomposition of a rooted balanced binary tree, 
our bound is $O(t\alpha(t) + n)$. 
%
%
%In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
%However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time of the algorithm of \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$ time.

We complement \Thm{main-alg} with a lower bound, suggesting that the comparison model complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
on some input.

\begin{theorem} 
\label{thm:main-lb}
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e., $d=2$) with the following properties.
Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
%
%the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
%the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

In other words, for any choice of $\sum_{p \in P} |p|\log |p|$, we can design a hard family where any algorithm
requires this time in the worst case, and our algorithm matches this bound.

\paragraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
characterize hard instances.
% explain \emph{when}
% computing contour trees is expensive, and provide insight into the hardness of particular instances. 
%It is intriguing
%that the cost (as defined earlier) roughly captures the hardness of an instance.
Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
This is tangentially related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. In general, our analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald \etal \cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour tree definitions and algorithms.

There are numerous algorithms for computing contour trees.
van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
This result forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
The latter result crucially uses monotone paths to improve the running time, and our algorithm repeatedly
exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
in $\MM$ and in its contour tree $\cC(\MM)$.

@@

We assume that $f$ has distinct values on all vertices, except for boundaries,
as expressed in the following constraint, which intuitively requires that 
all boundaries look like (local) horizontal cuts.

\begin{definition} \label{def:bound} The function $f$ is \emph{boundary critical} if the following holds.
Consider a boundary facet $F$. All vertices of $F$ have the same function value. Furthermore, all
neighbors of vertices in $F$, which are not also in $F$ itself, 
either have all function values strictly greater than or all function values strictly less than the function value at $F$.
\end{definition}

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, and this includes the infinite facet. 
We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time traversals of neighboring simplices
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$.

\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x| f(x) = h\}$.
A \emph{contour} is a maximal connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
\end{definition}

Note that a contour that does not contain a boundary is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
We let $\delta$ and $\eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
$f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.

@@

The set of critical points is denoted by $\cV(f)$.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called \emph{critical}, if $f(v) = h$, for some $v \in \cV(f)$. 
The critical points are exactly where the topology of level sets change.
By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maxima or minima.

\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
such that no $x \in p$ belongs to a critical contour.
\end{definition}

This equivalence relation gives a set \emph{contour classes}. By continuity, a regular point is present on a single contour,
and we can partition the regular contours into these classes. Every such class maps to intervals
of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
to be created at $x_i$ and destroyed at $x_j$.

\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV= \cV(f)$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher to lower function value.)
\end{definition}

@@

\subsection{Some technical remarks}
Note that if one intersects $\MM$ with given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $f$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degree will have $>1$ degree in $\reeb(\MM)$.

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explictly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

\section{A tour of the new contour tree algorithm} \label{sec:approach}

>>>>

\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}%
}
\date{}

@@

Let us interpret the running time. 
For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
A better upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
This is clearly an improvement when the tree is short and fat, but it can be even better. 
Consider a contour tree that is constant balanced, meaning that the ratio of sizes of sibling subtrees
is $\Theta(1)$. 
A calculation yields that for any decomposition of such a contour tree, 
our bound is $O(t\alpha(t) + n)$. 
%
%In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
%However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below,
so $t = \Omega(n)$. The running time of the algorithm of \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$ time.

We complement \Thm{main-alg} with a lower bound, suggesting that the comparison complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
on some input.

\begin{theorem} 
\label{thm:main-lb}
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
%
%the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
%the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

In other words, for any choice of $\sum_{p \in P} |p|\log |p|$, we can design a hard family where any algorithm
requires this cost in the worst case, and our algorithm matches this bound.

\paragraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
characterize hard instances.
% explain \emph{when}
% computing contour trees is expensive, and provide insight into the hardness of particular instances. 
%It is intriguing
%that the cost (as defined earlier) roughly captures the hardness of an instance.
Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
This is tangentially related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. From a mathematical perspective, this sort of analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald \etal \cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour trees definitions and algorithms.

There are numerous algorithms for computing contour trees.
van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
This result forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
The latter result crucially uses monotone paths to improve the running time, and we repeatedly
exploit numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
in $\MM$ and in its contour tree $\cC(\MM)$.

@@

We assume that $f$ has distinct values on all vertices, except for boundaries,
as expressed in the following constraint.

\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
Consider a boundary facet $F$. All vertices of $F$ have the same function value. Furthermore, all
neighbors of $F$ not in $F$ either have all function values strictly greater or strictly less
than this value.
\end{definition}

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, and this includes the infinite facet. 
We think of $d$ as constant, and assume that $\MM$ is represented in a data structure that allow constant-time traversals of neighboring simplices
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$.

\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x| f(x) = h\}$.
A contour is a maximal connected component of a level set. An $h$-contour is a contour where $f$-values are $h$.
\end{definition}

Note that a contour that does not contain a boundary is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
$f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.

@@

We use $\cV(f)$ to denote the set of critical points.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called critical, if $f(v) = h$, for some $v \in \cV(f)$. 
The critical points are exactly where the topology of level sets change.
By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maxima or minima.

\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
if the following holds. There exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
such that no $x \in p$ belongs to a critical contour.
\end{definition}

These give the \emph{contour classes}. By continuity, a regular point is present on a single contour,
and we can partition the regular contours into these classes. Every such class maps to intervals
of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
to be created at $x_i$ and destroyed at $x_j$.

\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher to lower function value.)
\end{definition}

@@

\subsection{Some technical remarks}

The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
contour in $f$. The up-degree (as opposed to \emph{Morse} up-degree) is defined as the number of $(f(x)+\delta)$-contours
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degree will have $>1$ degree in $\reeb(\MM)$.

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explictly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ will all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

\section{A tour of the contour tree algorithm} \label{sec:approach}
